### 2013-06-01 doc/dissertation/document/chapters/linear_problem/linear_problem.tex
adding data for estimators from hmcsa
updating spn chapter
making the linear problem chapter an appendix

\chapter{Conventional Solution Methods for Linear Systems}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems. In many cases, these
matrices are sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide the necessary background for the
remainder of this work. Details on the parallelization of conventional
methods are discussed.\footnote{The contents of this chapter,
  particularly those sections relating to projection methods and
  matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

\section{Preliminaries}
\label{sec:linear_preliminaries}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
be generated by inverting $\ve{A}$ either directly or indirectly:
\begin{equation}
  \ve{x} = \ve{A}^{-1} \ve{b}
  \label{eq:linear_problem_solution}\:.
\end{equation}
In addition we can define the residual:
\begin{equation}
  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
  \label{eq:linear_residual}
\end{equation}
such that an exact solution $\ve{x}$ has been found when
$\ve{r}=\ve{0}$.  From the statement in
Eq~(\ref{eq:linear_problem_solution}) we can already place a
restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
  meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
  will focus our efforts on approximately inverting the operator
  through various means.

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{Eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computable
from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003}
as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}
In addition, for problems that have a large scale over which the
independent variables may exist (e.g. a problem with events on
timescales ranging from nanoseconds to hours), a good measure of this
range is supplied by the \textit{stiffness ratio}:
\begin{equation}
  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
\end{equation}
Those problems that have a wide range of scales in their independent
variables, which will then be reflected in the operator, will then
have a large stiffness ratio. We will define such problems with large
stiffness ratios as \textit{stiff}.

General to both matrices and vectors, \textit{norms} are a mechanism
for collapsing objects of many elements to a single value. Per
LeVeque's text \citep{leveque_finite_2007}, the q-norm of a vector is defined
as:
\begin{equation}
  ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
  \mathbb{R}^N\:,\ q \in \mathbb{Z}^+
  \label{eq:q_norm}
\end{equation}
where ${v_i}$ is the $i^{th}$ component of the vector. Depending on
the value chosen for $q$, local or global qualities of the vector may
be obtained. For example, $q=2$ provides the root of a quadrature sum
of all elements in the vector giving a global measure of the vector
while $q=\infty$ gives the maximum value in the vector, a local
quantity that does not give information regarding the other elements
in the vector.

We can also compute the norm of a matrix by inferring from the norm of
the vector on which it is operating. Per LeVeque, we search for a
constant that is equivalent to $||\ve{A}||$:
\begin{equation}
  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
  \label{eq:matrix_norm_inequality}
\end{equation}
where the minimum value of $C$ that satisfies
Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
simple terms for common norms including the infinity norm:
\begin{equation}
  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
  \label{eq:matrix_infinity_norm}
\end{equation}
and the 2-norm:
\begin{equation}
  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
  \label{eq:matrix_2_norm}
\end{equation}
where $\rho$ is the spectral radius as defined in
Eq~(\ref{eq:spectral_radius}).

Knowing this, we can then define several useful properties of matrices
including the \textit{condition number}:
\begin{equation}
  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
  \label{eq:condition_number}
\end{equation}
which gives as a metric on assessing how close to singular the system
is. This is due to the fact $||\ve{A}^{-1}||$ is large near
singularities (and undefined for a singular matrix) and thus a large
condition number will be generated. We define such matrices as
\textit{ill-conditioned}.

@@

We can generate a more physical and geometric-based understanding of
these constraints by writing the new residual as $\tilde{\ve{r}} =
\ve{r}_0 - \ve{A}\boldsymbol{\delta}$ and again asserting the residual
must be orthogonal to $\mathcal{L}$. If $\tilde{\ve{r}}$ is to be
orthogonal to $\mathcal{L}$, then $\ve{A}\boldsymbol{\delta}$ must be
the projection of $\ve{r}_0$ onto the subspace $\mathcal{L}$ that
eliminates the components of the residual that exist in
$\mathcal{L}$. This situation is geometrically presented in
Figure~\ref{fig:linear_projection_constraint}.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.75}{
      \input{chapters/linear_problem/orthogonal_residual.pdftex_t} }
  \end{center}
  \caption{\textbf{Orthogonality constraint of the new residual with
      respect to $\mathcal{L}$.} \textit{By projecting $\ve{r}_0$ onto
      the constraint subspace, we minimize the new residual by
      removing those components.}}
  \label{fig:linear_projection_constraint}
\end{figure}

From Figure~\ref{fig:linear_projection_constraint} we then note that
the following geometric condition must hold:
\begin{equation}
  ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
  \mathbb{R}^N\:,
  \label{eq:linear_min_res}
\end{equation}
meaning that the residual of the system will always be
\textit{minimized} with respect to the constraints.

Given this minimization condition for the residual, we can form the
outline of an iterative projection method. Consider a matrix $\ve{V}$
to form a basis of $\mathcal{K}$ and a matrix $\ve{W}$ to form a basis
of $\mathcal{L}$. As $\boldsymbol{\delta} \in \mathcal{K}$ by
definition in Eq~(\ref{eq:linear_projection_step}), then
$\boldsymbol{\delta}$ can instead be rewritten as:
\begin{equation}
  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{R}^N:\,
  \label{eq:linear_delta_projection}
\end{equation}
where $\ve{V}$ \textit{projects} $\ve{y}$ onto $\mathcal{K}$. From the
orthogonality constraint in Eq~(\ref{eq:linear_projection_constraint})
it then follows that:
\begin{equation}
  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
  \label{eq:linear_constraint_projection}
\end{equation}
where here the projection onto $\mathcal{K}$ is constrained by the
projection onto $\mathcal{L}$. Knowing this, we can then outline the
following iteration scheme for a projection method:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}^k\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
  \end{gather}
  \label{eq:linear_projection_iteration}
\end{subequations}
where $\ve{V}$ and $\ve{W}$ are generated from the definitions of
$\mathcal{K}$ and $\mathcal{L}$ and are updated prior to each
iteration.

>>>>

\chapter{Conventional Solution Methods for Linear Systems}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems. In many cases, these
matrices are sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide the necessary background for the
remainder of this work. Details on the parallelization of conventional
methods are discussed.\footnote{The contents of this appendix,
  particularly those sections relating to projection methods and
  matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

@@

We can generate a more physical and geometric-based understanding of
these constraints by writing the new residual as $\tilde{\ve{r}} =
\ve{r}_0 - \ve{A}\boldsymbol{\delta}$ and again asserting the residual
must be orthogonal to $\mathcal{L}$. If $\tilde{\ve{r}}$ is to be
orthogonal to $\mathcal{L}$, then $\ve{A}\boldsymbol{\delta}$ must be
the projection of $\ve{r}_0$ onto the subspace $\mathcal{L}$ that
eliminates the components of the residual that exist in
$\mathcal{L}$. This situation is geometrically presented in
Figure~\ref{fig:linear_projection_constraint}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.75}{
      \input{chapters/linear_problem/orthogonal_residual.pdftex_t} }
  \end{center}
  \caption{\textbf{Orthogonality constraint of the new residual with
      respect to $\mathcal{L}$.} \textit{By projecting $\ve{r}_0$ onto
      the constraint subspace, we minimize the new residual by
      removing those components.}}
  \label{fig:linear_projection_constraint}
\end{figure}

From Figure~\ref{fig:linear_projection_constraint} we then note that
the following geometric condition must hold:
\begin{equation}
  ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
  \mathbb{R}^N\:,
  \label{eq:linear_min_res}
\end{equation}
meaning that the residual of the system will always be
\textit{minimized} with respect to the constraints.
Given this minimization condition for the residual, we can form the
outline of an iterative projection method. Consider a matrix $\ve{V}$
to form a basis of $\mathcal{K}$ and a matrix $\ve{W}$ to form a basis
of $\mathcal{L}$. As $\boldsymbol{\delta} \in \mathcal{K}$ by
definition in Eq~(\ref{eq:linear_projection_step}), then
$\boldsymbol{\delta}$ can instead be rewritten as:
\begin{equation}
  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{R}^N:\,
  \label{eq:linear_delta_projection}
\end{equation}
where $\ve{V}$ \textit{projects} $\ve{y}$ onto $\mathcal{K}$. From the
orthogonality constraint in Eq~(\ref{eq:linear_projection_constraint})
it then follows that:
\begin{equation}
  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
  \label{eq:linear_constraint_projection}
\end{equation}
where here the projection onto $\mathcal{K}$ is constrained by the
projection onto $\mathcal{L}$. Knowing this, we can then outline the
following iteration scheme for a projection method:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}^k\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
  \end{gather}
  \label{eq:linear_projection_iteration}
\end{subequations}
where $\ve{V}$ and $\ve{W}$ are generated from the definitions of
$\mathcal{K}$ and $\mathcal{L}$ and are updated prior to each
iteration.

### 2013-06-01 doc/dissertation/document/chapters/mc_background/mc_background.tex
additions to spn chapter
making the linear problem chapter an appendix

%%---------------------------------------------------------------------------%%
\section{Preliminaries}
\label{sec:mc_preliminaries}
We begin our discussion of Monte Carlo methods using these texts by
seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
operator $\ve{A}$, we can use diagonal splitting in a similar manner
as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
define the following operator\footnote{It should be noted that
  non-diagonal splittings have been recently explored
  \citep{srinivasan_monte_2010} and have the potential to improve
  efficiency. However, it was observed in this work that this type of
  splitting did not improve performance in the asymptotic limit of
  $\rho(\ve{H}) \rightarrow 1$ for non-trivial problems.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:,
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
such that we are solving the system:
\begin{equation}
  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
  \label{eq:richardson_split}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this Neumann series to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation by truncation will therefore lead
to an approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering the $i^{th}$
component of the solution vector:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
sequence permutation. We can generate these sequences of transitions
through Monte Carlo random walks by assigning them both a probability
and weight. As a reinterpretation of the iteration matrix, we then
form the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence and therefore truncate the Neumann series it is
approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
use the provided linear operator $\ve{A}$ to form the Neumann Ulam
decomposition while the adjoint method will use the adjoint linear
operator ($\ve{A}^T$ for real-valued systems) to form the
decomposition.

@@

\subsection{Adjoint Method: Evolution of a Solution}
\label{subsec:adjoint_evolution}
As a means of visually demonstrating the adjoint Monte Carlo method,
again consider the 2-dimensional thermal diffusion problem with
sources on the left and right hand sides of the domain and a smaller
uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
adjoint method with the collision estimator, the number of histories
sampled from the source was increased from 10 to 10,000,000 in order
to show its effects on the solution and the statistical nature of the
method. Figure~\ref{fig:adjoint_evolution} gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/mc_background/adjoint_evolution.png}
  \end{center}
  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 10
      histories per state. Top right: 1,000 histories per
      state. Bottom left: 100,000 histories per state. Bottom right:
      10,000,000 histories per state.}}
  \label{fig:adjoint_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tally
contributions are made. At 10,000,000 histories per state, enough
information has been tallied to generate a reasonable estimate for the
structure of the solution. The visual difference between
Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
precisely that determined by their mathematics. As the adjoint
solution evolves with the addition of histories, more histories
emanate from the boundary and smaller uniform source with more
penetrating from the boundary into the domain and making contributions
to the tallies in those states.
\clearpage

@@

@@

\subsection{Alternative Fixed Point Iterations}
\label{subsec:mcsa_fp_alternatives}

@@

%%---------------------------------------------------------------------------%%
\section{Variance Reduction Techniques}
\label{sec:variance_reduction}
Variance reduction is a mechanism by which the probabilities and
weights of the Monte Carlo problem are modified to improve the
performance of a given estimator. In many cases, such modifications
will introduce bias into the system. However, using Monte Carlo within
an MCSA iterative scheme provides a potential buffer for the solution
from this bias as the correction generated by the Monte Carlo solve
will contain large statistical error even without variance reduction.

>>>>

\section{Preliminaries}
\label{sec:linear_preliminaries}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
be generated by inverting $\ve{A}$ either directly or indirectly:
\begin{equation}
  \ve{x} = \ve{A}^{-1} \ve{b}
  \label{eq:linear_problem_solution}\:.
\end{equation}
In addition we can define the residual:
\begin{equation}
  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
  \label{eq:linear_residual}
\end{equation}
such that an exact solution $\ve{x}$ has been found when
$\ve{r}=\ve{0}$.  From the statement in
Eq~(\ref{eq:linear_problem_solution}) we can already place a
restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
will focus our efforts on approximately inverting the operator through
various means.

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computable from
the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral radius},
$\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003} as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}

%%---------------------------------------------------------------------------%%
\section{Neumann-Ulam Method}
\label{sec:mc_preliminaries}
We begin our discussion of Monte Carlo methods by seeking a solution
to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
we can use diagonal splitting in a similar manner as the stationary
method in Eq~(\ref{eq:linear_split_equation2}) to define the following
operator\footnote{It should be noted that non-diagonal splittings have
  been recently explored \citep{srinivasan_monte_2010} and have the
  potential to improve efficiency. However, it was observed in this
  work that this type of splitting did not improve performance in the
  asymptotic limit of $\rho(\ve{H}) \rightarrow 1$ for non-trivial
  problems.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:,
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
such that we are solving the system:
\begin{equation}
  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
  \label{eq:richardson_split}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this Neumann series to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation by truncation will therefore lead
to an approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering the $i^{th}$
component of the solution vector:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
sequence permutation. We can generate these sequences of transitions
through Monte Carlo random walks by assigning them both a probability
and weight. As a reinterpretation of the iteration matrix, we then
form the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence and therefore truncate the Neumann series it is
approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
use the provided linear operator $\ve{A}$ to form the Neumann Ulam
decomposition while the adjoint method will use the adjoint linear
operator ($\ve{A}^T$ for real-valued systems) to form the
decomposition.

@@

\subsection{Adjoint Method: Evolution of a Solution}
\label{subsec:adjoint_evolution}
As a means of visually demonstrating the adjoint Monte Carlo method,
again consider the 2-dimensional thermal diffusion problem with
sources on the left and right hand sides of the domain and a smaller
uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
adjoint method with the collision estimator, the number of histories
sampled from the source was increased from 10 to 10,000,000 in order
to show its effects on the solution and the statistical nature of the
method. Figure~\ref{fig:adjoint_evolution} gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/mc_background/adjoint_evolution.png}
  \end{center}
  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 10
      histories per state. Top right: 1,000 histories per
      state. Bottom left: 100,000 histories per state. Bottom right:
      10,000,000 histories per state.}}
  \label{fig:adjoint_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tally
contributions are made. At 10,000,000 histories per state, enough
information has been tallied to generate a reasonable estimate for the
structure of the solution. The visual difference between
Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
precisely that determined by their mathematics. As the adjoint
solution evolves with the addition of histories, more histories
emanate from the boundary and smaller uniform source with more
penetrating from the boundary into the domain and making contributions
to the tallies in those states.

@@

\subsubsection{Alternative Fixed Point Iterations}
\label{subsubsec:alternative_fixed_point}

@@

@@

%%---------------------------------------------------------------------------%%
\section{Variance Reduction Techniques}
\label{sec:variance_reduction}
Variance reduction is a mechanism by which the probabilities and
weights of the Monte Carlo problem are modified to improve the
performance of a given estimator. In many cases, such modifications
will introduce bias into the system. However, using Monte Carlo within
an MCSA iterative scheme provides a potential buffer for the solution
from this bias as the correction generated by the Monte Carlo solve
will contain large statistical error even without variance
reduction. For this work, any method that alters the weights and
probabilities of the Monte Carlo method with the aim of improving
either iterative performance or time to convergence will be defined as
a variance reduction scheme.
